Non-word repetition in Yélî Dnye

### 2021-02-11 manuscript.Rmd
proofing

---
title             : "Non-word repetition in Yélî Dnye"
shorttitle        : "NWR in Yélî Dnye"

@@

affiliation:
  - id            : "1"
    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University"
  - id            : "2"
    institution   : "Max Planck Institute for Psycholinguistics"
  - id            : "3"
    institution   : "University of Chicago"

abstract: >
   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development at a global and individual level via three contributions. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.  Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.

keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
wordcount         : "~11,00 words"

@@

@@

@@

d2r=function(d,n1,n2) {
  a=(n1+n2)^2/(n1*n2)
  d/sqrt(d^2 +a)}
```

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas.

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly jittered for ease of inspection of different lengths at a given age."}

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

<!--M2A: We mention "three" contributions in the abstract but then say above "The study was designed to contribute to four broad research areas." and add this as "fourth" in our Introduction. This is minor, but feels slightly inconsistent. Happy for you to handle it as you most prefer!-->
Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge). There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance, better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have (unconditioned) contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.

@@

We could include `r summary(xl_fr)$devcomp$dims["N"]` observations, from `r summary(xl_fr)$ngrps["file"]` children producing in any given trial one of `r summary(xl_fr)$ngrps["target"]` potential target words. The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std. Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq). Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.

@@

We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.  Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`]. Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r round(cor_ty_nwr$estimate,3)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`. Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic phone frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus phone frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.

@@

For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]). These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English. This result suggests that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.] 
The complex consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("complex")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("complex")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("complex")],na.rm=T)`%); 
whereas simple consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("simple")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("simple")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("simple")],na.rm=T)`%). 
Again these groups of sounds differ in their frequency within the language. Their type frequencies in Yélî Dnye are: complex consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰) versus simple consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰).

Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`). Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.

>>>>

---
title             : "Non-word repetition in children learning Yélî Dnye"
shorttitle        : "NWR in Yélî Dnye learners"

@@

affiliation:
  - id            : "1"
    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University"
  - id            : "2"
    institution   : "Max Planck Institute for Psycholinguistics"
  - id            : "3"
    institution   : "University of Chicago"

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
wordcount         : "~11,500 words"

@@

# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)

@@

printr=function(x) sub("0.", ".", round(x,3))

@@

d2r=function(d,n1,n2) {
  a=(n1+n2)^2/(n1*n2)
  d/sqrt(d^2 +a)}

sink("lastknit_session_info.txt")
sessionInfo()
sink()
```

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001non-word], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006non-word], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006non-word]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly shifted for ease of inspection of different lengths at a given age."}

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021non-word]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014non-word; @vance2005speech]. Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge]. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations). Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.

Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014non-word], with the entire test inventory of 40 final items distributed across children.

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.

@@

We could include `r summary(xl_fr)$devcomp$dims["N"]` observations, from `r summary(xl_fr)$ngrps["file"]` children producing in any given trial one of `r summary(xl_fr)$ngrps["target"]` potential target words. The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std. Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq). Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r printr(xlfr_r$estimate)`.

@@

We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.  Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`]. Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r printr(cor_ty_nwr$estimate)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`. Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age. As before, the model contained random slopes for both child ID and target. In this model, both cross-linguistic phone frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std. Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus phone frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std. Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.

@@

For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]). These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.] 
The complex consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("complex")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("complex")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("complex")],na.rm=T)`%); 
whereas simple consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("simple")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("simple")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("simple")],na.rm=T)`%). 
Again these groups of sounds differ in their frequency within the language. Their type frequencies in Yélî Dnye are: complex consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰) versus simple consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰).

Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`%). Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
